---
title: "Optimisation Prac 2"
author: "Andomei Smit: SMTAND051"
date: "25/02/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

# Simulated Annealing

## Changing the solution to the travelling salesman problem
 In this case we will change the way that the moves are made to change the order of the cities

```{r}
# load data
y <- as.matrix(eurodist)
```


```{r}
# function to initialise cities
get_initial_x <- function(ncity){
  tour <- sample(1:ncity)
  # ensure tour returns to first city
  tour <- c(tour, tour[1])
  
  return(tour)
}

# get initial tour:
cur_tour <- get_initial_x(nrow(y))
```


```{r}
# define an evaluation function
evaluate_x <- function(dists, tour){
  sum(dists[cbind(tour[-length(tour)], tour[-1])])
}
```

We will change the function that perturbs by setting it to choose 2 random indeces and swopping the cities related to those cities.

```{r}
perturb_x <- function(tour){
  
  # select two indeces at random
  i_j <- sample(1:(length(tour)), 2, replace = F)
  # swop the cities at these indeces
  ## select city i
  city_i <- tour[i_j[1]]
  ## select city j
  city_j <- tour[i_j[2]]
  ## swop them
  tour[i_j[1]] <- city_j
  tour[i_j[2]] <- city_i
  
  return(tour)
}

```

```{r}
# set start temperature and geometric cooling factor
set.seed(100) # for repeatability
start_temp <- 50000
temp_factor <- 0.98 
```


```{r}
# get an initial solution
cur_x <- get_initial_x(ncol(y))
# evaluate the solution
cur_fx <- evaluate_x(dists = y, tour = cur_x)
```


```{r}
# initialize results data frames
all_fx <- c()
all_x <- data.frame()
# for a fixed number of iterations
for(i in 1:10000){
  
  # generate a candidate solution
  prop_x <- perturb_x(cur_x)
  
  # evaluate the candidate solution
  prop_fx <- evaluate_x(dists = y, tour = prop_x)
  
  # calculate the probability of accepting the candidate
  anneal_temp <- start_temp * temp_factor ^ i
  accept_prob <- exp(-(prop_fx - cur_fx) / anneal_temp)
  
  # accept or reject the candidate
  if(runif(1) < accept_prob){
   cur_x <- prop_x
   cur_fx <- prop_fx
  }
  
  # store all results
  all_fx <- c(all_fx, cur_fx)
  all_x <- rbind(all_x,cur_x)
  
}
```


```{r}
plot(all_fx,type="l")
```

The best tour returned by the simulated annealing algorithm is given below.


```{r}
best_solution <- which.min(all_fx)
best_tour <- all_x[best_solution,]
best_solution
best_tour
colnames(y)[as.numeric(best_tour)]
```

## Changing the solution to the Soduko problem

```{r}
# create a Sudoku puzzle 
s <- matrix(0,ncol=9,nrow=9)
s[1,c(6,8)] <- c(6,4)
s[2,c(1:3,8)] <- c(2,7,9,5)
s[3,c(2,4,9)] <- c(5,8,2)
s[4,3:4] <- c(2,6)
s[6,c(3,5,7:9)] <- c(1,9,6,7,3)
s[7,c(1,3:4,7)] <- c(8,5,2,4)
s[8,c(1,8:9)] <- c(3,8,5)
s[9,c(1,7,9)] <- c(6,9,1)

s
```



```{r}
# define the spaces that can be changed
free_spaces <- (s == 0)
free_spaces
```


```{r}
get_initial_x <- function(s){
  
  # identify free spaces
  free_spaces <- (s == 0)
  # fixed non-free spaces
  cur_x <- s
  # randomly choose a number between 1 and 9 for free spaces
  cur_x[free_spaces] <- sample(1:9, sum(free_spaces), replace=T)
  
  return(cur_x)
}

# create the initial solution:
init_x <- get_initial_x(s)
```

These functions will stay the same.

```{r}
# function to count duplicates:
count_duplicates <- function(x)
{
  n_dup <- length(x) - length(unique(x))
  return(n_dup)
}

# sum duplicates accross rows, columns and blocks:
evaluate_x <- function(x){
  
  # within-row duplications
  row_dups <- sum(apply(x,1,count_duplicates))
  
  # within-col duplications
  col_dups <- sum(apply(x,2,count_duplicates))
  
  # within-block duplications
  block_dups <- 0
  for(i in 1:3){
    for(j in 1:3){
      small_x <- x[(3*(i-1) + 1):(3*i), (3*(j-1) + 1):(3*j)]
      thisblock_dups <- count_duplicates(as.vector(small_x))
      block_dups <- block_dups + thisblock_dups
    }
  }
  
  total_dups <- row_dups + col_dups + block_dups
  
  return(total_dups)
  
} 

# evaluate performance of the initial solution:
evaluate_x(init_x)
```


We will change the perturbing in this problem to replace one of the free_spaces values in the column and row that has the most duplicates with a random draw. The idea is that if we target the most problematic columns first, the algorithm should run faster.

```{r}
perturb_x <- function(x, free_spaces)
{
  # find column with the largest number of duplicates
  max_col_index <- which.max(apply(init_x, 2, count_duplicates))
  
  # select a free site
  row_index <- sample(1:9,1,prob=free_spaces[,max_col_index])
  # change that site at random
  x[row_index, max_col_index] <- sample(1:9,1,replace=T)
  
   # find row with the largest number of duplicates
  max_row_index <- which.max(apply(init_x, 1, count_duplicates))
  
  # select a free site
  col_index <- sample(1:9,1,prob=free_spaces[max_row_index,])
  
  # change that site at random
  x[max_row_index, col_index] <- sample(1:9,1,replace=T)
  
  return(x)
}
```

We now apply simulated annealing to solve the Sudoku puzzle, starting by setting a start temperature and geometric cooling factor.

```{r}
set.seed(100) # for repeatability
start_temp <- 1e3
temp_factor <- 0.995 
```

We'll generate a new initial solution and evaluate, just to keep this whole code block self-contained.

```{r}
cur_x <- get_initial_x(s)
cur_fx <- evaluate_x(cur_x)
```

We now apply the simulated annealing algorithm.

```{r}
# initial results data frames
all_fx <- c()
all_x <- data.frame()
# for a fixed number of iterations
for(i in 1:3000){ 
  
  # generate a candidate solution
  prop_x <- perturb_x(cur_x, free_spaces = free_spaces)
  
  # evaluate the candidate solution
  prop_fx <- evaluate_x(prop_x)
  
  # calculate the probability of accepting the candidate
  anneal_temp <- start_temp * temp_factor ^ i
  accept_prob <- exp(-(prop_fx - cur_fx) / anneal_temp)
  
  # accept or reject the candidate
  if(runif(1) < accept_prob){
    cur_x <- prop_x
    cur_fx <- prop_fx
  }
  
  # store all results
  all_fx <- c(all_fx, cur_fx)
  all_x <- rbind(all_x,as.vector(cur_x))
  
}
```

We plot the solution quality over time. 


```{r}
plot(all_fx,type="l")
```
From the above plot we can see that this peturbing function actually performs worse than the previous model.


